---
title: "EDA_MT"
author: "Kira Dahinden"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

import
```{r}
library(dplyr)
library(ggplot2)
library(lattice)
library(gridExtra)
library(ggpubr)
library(mgcv)
library(tibble)
library(magrittr)
```


#GRAPHICAL ANALYSIS

## Getting data
```{r}
# Lese die RDS-Datei ein
rawdata <- readRDS("/Users/kira/Desktop/vectomax_ready.RDS")
```

## Conversions of various data:

- rawdata
- model_data (wegen welchem modell?gamlss?)
- model_data.df (gammSlice)
- model_data_scaled (gammSclice)
- (dummy??)


### model_data
```{r Create model_data}
model_data <- rawdata[c("Aa.adults", "weeks.since.treatment", "yday_end", "region", "site", "manhole")]
model_data <- na.omit(model_data)

```

```{r}
#compared to rawdata (for weeks since treatment)
unique(rawdata$weeks.since.treatment) 
unique(model_data$weeks.since.treatment) # same 
table(rawdata$weeks.since.treatment)
table(model_data$weeks.since.treatment)
```
```{r}
#for yday_end
unique(rawdata$yday_end) 
unique(model_data$yday_end) # same amount
table(rawdata$yday_end)
table(model_data$yday_end)
```

```{r}
#for manhole
unique(rawdata$manhole) 
table(rawdata$manhole)

```


```{r}
unique(model_data$manhole) # still 20
table(model_data$manhole)
```


### model_data.df 
Preparation for gammSlice: make data into df and scale! 
```{r}
# gammslice does not support tibbles! Therefore, save as a data frame
# -> Note: Other models do not require it as a data frame
model_data.df <- as.data.frame(model_data)
```

```{r}
unique(model_data.df$weeks.since.treatment) 
table(model_data.df$weeks.since.treatment)
```

#### model_data_scaled:
```{r}
# The scale() function in R standardizes numeric variables by transforming them to a scale with a mean of 0 and a standard deviation of 1 (standardization):
# Standardize weeks.since.treatment and yday_end
# Create a copy of the original DataFrame
model_data_scaled <- model_data.df

# Standardize the desired variables and save them in the new copy
model_data_scaled$weeks.since.treatment <- scale(model_data.df$weeks.since.treatment)
model_data_scaled$yday_end <- scale(model_data.df$yday_end)
```

Still 21 values, BUT they are now scaled relative to the standard deviation of the data
```{r}
#unique(model_data_scaled$weeks.since.treatment) 
#table(model_data_scaled$weeks.since.treatment)
```



```{r}
# Check dimensions: reduced from 43 to 6 because we only need 6
dim(rawdata)
dim(model_data)
dim(model_data.df)
dim(model_data_scaled)
```

```{r}
# structure & head
str(rawdata)
head(rawdata)

```

```{r}
str(model_data)
head(model_data)
```

```{r}
str(model_data.df)
head(model_data.df)
```

```{r}
#Important: Factors such as region, site, and manhole were not altered, as they do not need to be scaled
str(model_data_scaled)
head(model_data_scaled)
```


```{r}
# Number of missing values per column
colSums(is.na(rawdata))
```

```{r}
colSums(is.na(model_data))
```

```{r}
colSums(is.na(model_data.df))
```

```{r}
colSums(is.na(model_data_scaled))
```


```{r}
# Correlation between varibales before and AFTER scaling = same correlation
cor_original <- cor(model_data[, sapply(model_data, is.numeric)], use = "complete.obs")
cor_scaled <- cor(model_data_scaled[, sapply(model_data_scaled, is.numeric)], use = "complete.obs")

# compare correlation matricex
print(cor_original)
print(cor_scaled)

```
## Compare with the original mgcv model
```{r}
#For gammslice, scaling is necessary, so compare: (with the original model)

library(mgcv)

# Model with original Data
mod_mgcv_rawdat <- gam(Aa.adults ~ s(weeks.since.treatment) + s(yday_end), data = rawdata, family = "poisson")

```

```{r}
# Model with model_data 
mod_mgcv_modeldat <- gam(Aa.adults ~ s(weeks.since.treatment) + s(yday_end), data = model_data, family = "poisson")
```

```{r}
# Model wit hscaled data
mod_mgcv_scaleddat <- gam(Aa.adults ~ s(weeks.since.treatment) + s(yday_end), data = model_data_scaled, family = "poisson")
```

```{r}
summary(mod_mgcv_rawdat)
```

```{r}
summary(mod_mgcv_modeldat)
```

```{r}
summary(mod_mgcv_scaleddat)
```
--> Scaling had no effect on the statistical model fit.

# GRAPHICAL ANALYSIS 

Distribution of the response variables
We first create a proto-density plot. Note the square-root-scale used for the x-axis.
```{r}
library(ggplot2)

gg.dens <- ggplot(data = rawdata) +
  geom_density() +
  geom_rug(alpha = 0.1)+
  scale_x_sqrt()
```

We first visualise the distribution of the Aedes albopictus (“Aa”).
```{r}
gg.dens + aes(x = Aa.adults)
```


Now with model_data_scaled (for safety, but it should be the same since it's not scaled!)
```{r}

gg.dens <- ggplot(data = model_data_scaled) +
  geom_density() +
  geom_rug(alpha = 0.1)+
  scale_x_sqrt()
```

We first visualise the distribution of the Aedes albopictus (“Aa”).
```{r}
gg.dens + aes(x = Aa.adults)
```


```{r}
# Check the distribution of the variables before and after the transformation (since they are being transformed!).
ggplot(model_data, aes(x = weeks.since.treatment)) +
  geom_histogram(binwidth = 1) +
  ggtitle("Verteilung von weeks.since.treatment in model_data")

ggplot(model_data_scaled, aes(x = weeks.since.treatment)) +
  geom_histogram(binwidth = 0.5) +
  ggtitle("Verteilung von weeks.since.treatment in model_data_scaled")

```

Protoplot Aa (P.7)
```{r}
gg.Aa <- ggplot(data = rawdata,
                 mapping = aes(y = Aa.adults,
                               x = weeks.since.treatment,
                              group = manhole)) +
geom_point(alpha = 0.5) +
geom_line(alpha = 0.5) +
scale_y_sqrt()

#plot the adults of the “Aa” counts against time.
gg.Aa +
  geom_smooth(group = 1)
```
```{r}
gg.Aa.model <- ggplot(data = model_data,
                 mapping = aes(y = Aa.adults,
                               x = weeks.since.treatment,
                              group = manhole)) +
geom_point(alpha = 0.5) +
geom_line(alpha = 0.5) +
scale_y_sqrt()

gg.Aa.model +
  geom_smooth(group = 1)
```

Zoom into sites
```{r}
gg.Aa +
  facet_wrap(~site)
```


model_data
```{r}
gg.Aa.model <- ggplot(data = model_data,
                 mapping = aes(y = Aa.adults,
                               x = weeks.since.treatment,
                              group = manhole)) +
geom_point(alpha = 0.5) +
geom_line(alpha = 0.5) +
scale_y_sqrt()

gg.Aa.model +
  geom_smooth(group = 1)
```
Let’s zoom into sites as well.
```{r}
gg.Aa.model +
  facet_wrap(~site)
```


scaled 
```{r}
gg.Aa.scaled <- ggplot(data = model_data_scaled,
                 mapping = aes(y = Aa.adults,
                               x = weeks.since.treatment,
                              group = manhole)) +
geom_point(alpha = 0.5) +
geom_line(alpha = 0.5) +
scale_y_sqrt()

gg.Aa.scaled +
  geom_smooth(group = 1)
```
zoom into sites
```{r}
gg.Aa.scaled +
  facet_wrap(~site)
```

now focus on single manholes:
```{r}
library(scales)
gg.Aa +
  facet_wrap(~manhole) +
  rotate_x_text()
```

```{r}
gg.Aa.model +
  facet_wrap(~manhole) +
  rotate_x_text()
```

```{r}
gg.Aa.scaled +
  facet_wrap(~manhole) +
  rotate_x_text()
```

There are important differences among manholes. This may be due to: 1) local conditions or 2) the treatment applied to the each single manhole.




 







